Wide binaries in Gaia DR1

### 2016-09-28 paper/ms.tex
math

% define macros for math
\newcommand{\given}{\,|\,}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transp}[1]{{#1}^{\!\mathsf{T}}}
\newcommand{\inv}[1]{{#1}^{-1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\vperp}{\bs{v}^\perp}
\newcommand{\propm}{\bs{\mu}}
\newcommand{\matrx}[1]{\mathbf{#1}}
\newcommand{\kms}{\rm km~s^{-1}}

@@

In what follows, all vectors are column vectors.
Properties of the squared exponential:
\begin{eqnarray}
  \ln\left[\int\exp(-\frac{1}{2}\,[x-\mu]\transp\,A\inv\,[x-\mu] - \Delta)\,\dd x\right]
  &=& -\frac{d}{2}\,\ln 2\pi -\frac{1}{2}\ln ||A|| -\Delta
  \quad ,
\end{eqnarray}
where $x$ and $\mu$ are $d$-dimensional vectors, $A$ is a symmetric
matrix, $\Delta$ is a scalar, and the integral is implicitly over all
of $d$-space.
We will need to complete squares.
If we equate
\begin{eqnarray}
  \frac{1}{2}\,[x-\mu]\transp\,A\inv\,[x-\mu] + \Delta
  &=& \frac{1}{2}\,x\transp\,A\inv\,x + x\transp\,B\,b + C
  \quad ,
\end{eqnarray}
where $B\,b$ is a $d$-vector, and $C$ is a scalar, then we find
\begin{eqnarray}
  \mu &=& A\,B\,b
  \\
  C & = & \Delta - \frac{1}{2}\,\mu\transp\,A\inv\,\mu
  \quad .
\end{eqnarray}

Now in this case, we want to write down the probability of the

>>>>

% define macros for math
\newcommand{\given}{\,|\,}
\newcommand{\normal}{N}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transp}[1]{{#1}^{\!\mathsf{T}}}
\newcommand{\inv}[1]{{#1}^{-1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\vperp}{\bs{v}^\perp}
\newcommand{\propm}{\bs{\mu}}
\newcommand{\matrx}[1]{\mathbf{#1}}
\newcommand{\kms}{\rm km~s^{-1}}
\newcommand{\data}{\mathrm{data}}

@@

In what follows, all vectors are column vectors.
We indicate whether a symbol represents a scalar, vector, or tensor
(matrix) by context, not typesetting.
Properties of the squared exponential:
\begin{eqnarray}
  \ln\left[\int\exp(-\frac{1}{2}\,\transp{[x-\mu]}\,\inv{A}\,[x-\mu] - \Delta)\,\dd x\right]
  &=& -\frac{d}{2}\,\ln 2\pi -\frac{1}{2}\ln ||A|| -\Delta
  \quad ,
\end{eqnarray}
where $x$ and $\mu$ are $d$-dimensional vectors, $A$ is a symmetric
matrix, $\Delta$ is a scalar, and the integral is implicitly over all
of $d$-space.
We will need to complete squares.
If we equate
\begin{eqnarray}
  \frac{1}{2}\,\transp{[x-\mu]}\,\inv{A}\,[x-\mu] + \Delta
  &=& \frac{1}{2}\,\transp{x}\,\inv{A}\,x + \transp{x}\,B\,b + C
  \quad ,
\end{eqnarray}
where $B\,b$ is a $d$-vector, and $C$ is a scalar, then we find
\begin{eqnarray}
  \mu &=& A\,B\,b
  \\
  C & = & \Delta - \frac{1}{2}\,\transp{\mu}\,\inv{A}\,\mu
  \quad .
\end{eqnarray}

Now in this case, we want to write down the probability of the data
given the distance and velocity (the likelihood), and the prior pdf for
the velocity, and their product.
The prior pdf for the velocity is Gaussian:
\begin{eqnarray}
  p(v) &=& \normal(v\given 0,V)
  \\
  \ln p(v) &=& -\frac{3}{2}\,\ln\pi -\frac{1}{2}\,\ln||V|| -\frac{1}{2}\,\transp{v}\,\inv{V}\,v
  \\
  V &=& \sigma_v^2\,I
  \quad ,
\end{eqnarray}
where $v$ is the three-vector velocity, and $V$ is an isotropic
variance tensor.
Similarly, the likelihood for the data is also a Gaussian.
Before we write that down, we will make some definitions.
We construct a vector $y$ as follows:
\begin{eqnarray}
  \transp{y} &\equiv& [d_1\,\varpi_1 - 1, d_1\,\mu_{\alpha,1}, d_1\,\mu_{\delta,1}
                       d_1\,\varpi_2 - 1, d_2\,\mu_{\alpha,2}, d_1\,\mu_{\delta,2}]
  \\
  \transp{y} &\equiv& [d_1\,\varpi_1 - 1, d_1\,\mu_{\alpha,1}, d_1\,\mu_{\delta,1}, v_{r,1}
                       d_1\,\varpi_2 - 1, d_2\,\mu_{\alpha,2}, d_1\,\mu_{\delta,2}, v_{r,2}
  \quad ,
\end{eqnarray}
where we have given two versions, a 6-component and an 8-component
version (depending on whether there are measured radial velocities),
the subscripts 1, 2 refer to the first and second stars (arbitrarily
ordered) of the pair, we have multiplied the strict observables by
distances $d_1, d_2$, which we are permitted to do (because we are
\emph{conditioning} on $d_1, d_2$). The vector $y$ is a column vector
that we have transposed to make it easy to write down.  Fundamentally,
our model is
\begin{eqnarray}
  y &=& M\,v + \mathrm{noise}
  \\
  \transp{M} &=& [0, \hat{\alpha_1}, \hat{\delta_1},
                  0, \hat{\alpha_2}, \hat{\delta_2}]
  \\
  \transp{M} &=& [0, \hat{\alpha}_1, \hat{\delta}_2, \hat{r}_1
                  0, \hat{\alpha}_2, \hat{\delta}_2, \hat{r}_2]
  \quad,
\end{eqnarray}
where $M$ is a $6\times 3$ or $8\times 3$ design matrix, with zeros
in two rows and unit vectors along the other rows.
The noise is drawn from a Gaussian with ($6\times 6$ or $8\times 8$)
covariance matrix $C$, constructed from the distances $d_1, d_2$ and
the \gaia-provided individual-object covariance matrices.
Given all these definitions, the likelihood function is
\begin{eqnarray}
  p(\data\given v,d_1,d_2) &=& d_1^3\,d_2^3\,\normal(y\given M\,v, C)
  \\
  \ln p(\data\given v,d_1,d_2) &=& 3\,\ln d_1 + 3\,\ln d_2 -\frac{6~\mathrm{or}~8}{2}\,\ln\pi
  -\frac{1}{2}\,\ln||C|| -\frac{1}{2}\,\transp{[y-M\,v]}\,\inv{C}\,[y-M\,v]
  \quad ,
\end{eqnarray}
where the factor of $d_1^2\,d_2^3$ converts to units of \gaia\ data
from units of $y$.
